---
title: "Bus Count Analysis"
output: html_document
date: "2024-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(lubridate)
library(ggplot2)
```

```{r, include=FALSE}
# APC data
jan_APC <- read_xlsx('January.xlsx', sheet = 'clean_passengerCountDetailDump_', )
feb_APC <- read_xlsx('February.xlsx', sheet = 'passengerCountDetailDump_Feb202')
mar_APC <- read_xlsx('March.xlsx', sheet = 'passengerCountDetailDump_Mar202')

# Driver estimations data
jan_DRE <- read_xlsx('January.xlsx', sheet = 'Driver Estimations')
feb_DRE <- read_xlsx('February.xlsx', sheet = 'Driver Estimations')
mar_DRE <- read_xlsx('March.xlsx', sheet = 'Driver Estimations')
  
# Trip codes
trips <- read.csv('trips.csv')
```

## Clean driver estimation data

Some N/A observations are valid, such as inbound only or tripper trips, but they're complicating this process unnecessarily so they're going to be removed.

```{r}
# Filter out correct buses, remove non-regular lines, remove N/A observations
jan_DRE <- jan_DRE %>%
  filter(Vehicle %in% c(4025, 4027, 4028, 4029, 4030, 4031)) %>%
  filter(grepl("Regular", Line...11, fixed = TRUE)) %>%
  filter(!is.na(In) & !is.na(Out))
feb_DRE <- feb_DRE %>%
  filter(Vehicle %in% c(4027, 4030, 4031)) %>%
  filter(grepl("Regular", Line...11, fixed = TRUE)) %>%
  filter(!is.na(In) & !is.na(Out))
mar_DRE <- mar_DRE %>%
  filter(Vehicle %in% c(4027, 4031)) %>%
  filter(grepl("Regular", Line...11, fixed = TRUE)) %>%
  filter(!is.na(In) & !is.na(Out))

# Format datetime data as simple times
jan_DRE$`Start Time (Scheduled)` <- format(as.POSIXct(jan_DRE$`Start Time (Scheduled)`), format = '%H%M')
jan_DRE$`End Time (Scheduled)` <- format(as.POSIXct(jan_DRE$`End Time (Scheduled)`), format = '%H%M')

feb_DRE$`Start Time (Scheduled)` <- format(as.POSIXct(feb_DRE$`Start Time (Scheduled)`), format = '%H%M')
feb_DRE$`End Time (Scheduled)` <- format(as.POSIXct(feb_DRE$`End Time (Scheduled)`), format = '%H%M')

mar_DRE$`Start Time (Scheduled)` <- format(as.POSIXct(mar_DRE$`Start Time (Scheduled)`), format = '%H%M')
mar_DRE$`End Time (Scheduled)` <- format(as.POSIXct(mar_DRE$`End Time (Scheduled)`), format = '%H%M')
```

## Split driver estimation data to inbound and outbound trips

```{r}
split_trips <- function(data) {
  output <- data.frame(matrix(ncol = 25, nrow = 0))
  colnames(output) <- c("Vehicle","Date","Time","Pay Time","Direction","WC",
                    "Passengers","Line...11","Column","Line...13","Total",
                    "Miles","Veh Hrs...16","Adj Arr","Adj Dep","Trips",
                    "Late_Early","TravTime","CalcTime1","CalcTime2",
                    "Veh Hrs...24","Capac","Crowd","Crwd#","DNR")
  
  for (i in 1:nrow(data)) {
    # Gather existing info
    new_row <- data[i, !(colnames(data) %in% c('Start Time (Scheduled)', 'End Time (Scheduled)', 'In', 'Out'))]
    
    # P or Q
    if (data$Line...13[i] == 'P' | data$Line...13[i] == 'Q') {
      new_row$Time <- data$`Start Time (Scheduled)`[i]
      new_row$Passengers <- data$Out[i]
      new_row$Direction <- 'outbound'
      output <- rbind(output, new_row)
      
      new_row$Passengers <- data$In[i]
      new_row$Direction <- 'inbound'
      output <- rbind(output, new_row)
    }
    
    # Outbound
    else if (!is.na(data$Out[i])) {
      new_row$Time <- data$`Start Time (Scheduled)`[i]
      new_row$Passengers <- data$Out[i]
      new_row$Direction <- 'outbound'
      output <- rbind(output, new_row)
    }
    
    # Inbound
    else if (!is.na(data$In[i])) {
      new_row$Time <- data$`End Time (Scheduled)`[i]
      new_row$Passengers <- data$In[i]
      new_row$Direction <- 'inbound'
      output <- rbind(output, new_row)
    }
  }
  
  return(output)
}

# Split up trips and update dataset
jan_DRE <- split_trips(jan_DRE)
feb_DRE <- split_trips(feb_DRE)
mar_DRE <- split_trips(mar_DRE)
```

## Attribute trips to driver estimations

``` {r}
# Making trip codes
trip_code <- function(row) {
  
  # Search for day of week and initialize service code for it
  if (wday(row$Date) == 4) {
    service_id <- 3
  } else if (wday(row$Date) == 6) {
    service_id <- 5
  } else if (wday(row$Date) == 1 | wday(row$Date) == 7) {
    service_id <- 10
  } else {
    service_id <- 1
  }

  # P or Q
  if (row$Line...13 == 'P' | row$Line...13 == 'Q') {
    prelim_code <- paste(row$Line...13,
                         service_id,
                         row$Time,
                         sep = '_')
  }
  
  # Any other route
  else {
    prelim_code <- paste(row$Line...13,
                         service_id,
                         row$Direction,
                         row$Time,
                         sep = '_')
  }
  
  return(prelim_code)
}

# Assigning trip codes
find_trip <- function(data) {
  trip_return <- c()
  for (i in 1:nrow(data)) {
    # For winter break trips
    if (data$Date[i] >= as.POSIXct("2024-01-01",tz = 'UTC')
        &
        data$Date[i] <= as.POSIXct("2024-01-05",tz = 'UTC')) {
      trip_return <- c(trip_return, sub('_','_2',trip_code(data[i,])))
    }

    # For finals trips
    else if (data$Date[i] >= as.POSIXct("2024-03-18",tz = 'UTC')
             &
             data$Date[i] <= as.POSIXct("2024-03-22",tz = 'UTC')) {
      trip_return <- c(trip_return, sub('_','_1',trip_code(data[i,])))
    }

    # For spring break trips
    else if (data$Date[i] >= as.POSIXct("2024-03-23",tz = 'UTC')
             &
             data$Date[i] <= as.POSIXct("2024-03-31",tz = 'UTC')) {
      trip_return <- c(trip_return, sub('_','_2',trip_code(data[i,])))
    }

    # For all others
    else {
      trip_return <- c(trip_return, trip_code(data[i,]))
    }
  }
  return(trip_return)
}

# Assign the codes!
jan_DRE$trip_id <- find_trip(jan_DRE)
feb_DRE$trip_id <- find_trip(feb_DRE)
mar_DRE$trip_id <- find_trip(mar_DRE)
```

## Clean APC data and calculate passenger loads
```{r}
# Remove rows where there is no trip ID
jan_APC <- filter(jan_APC, !is.na(Trip))
feb_APC <- filter(feb_APC, !is.na(Trip))
mar_APC <- filter(mar_APC, !is.na(Trip))

# Calculate passenger loads
passenger_load <- function(data) {
  output <- data.frame(matrix(ncol = 11, nrow = 0))
  colnames(output) <- c("# Date","Time","Vehicle","Stop","Stop ID","Route",
                        "Direction","Trip","Boardings","Deboardings","Load")
  
  data$Load <- NA
  for (date in unique(data$`# Date`)) {
    date_subset <- data[data$`# Date` == date,]
    for (vehicle in unique(date_subset$Vehicle)) {
      vehicle_subset <- date_subset[date_subset$Vehicle == vehicle,]
      for (trip in unique(vehicle_subset$Trip)) {
        trip_subset <- vehicle_subset[vehicle_subset$Trip == trip,]
        trip_subset <- trip_subset[order(as.Date(trip_subset$Time, format="%m/%d/%Y")),]
        for (i in 1:nrow(trip_subset)) {
          if (i == 1) {
            trip_subset$Load[i] <- trip_subset$Boardings[i] - trip_subset$Deboardings[i]
          } else {
            trip_subset$Load[i] <- trip_subset$Load[i-1] + trip_subset$Boardings[i] - trip_subset$Deboardings[i]
          }
        }
        output <- rbind(output, trip_subset)
      }
    }
  }
  return(output)
}

# Calculate passenger loads
jan_APC <- passenger_load(jan_APC)
feb_APC <- passenger_load(feb_APC)
mar_APC <- passenger_load(mar_APC)
```

``` {r}
# Sum boardings and alightings and find max load grouped by date, vehicle, and trip code
jan_APC.agg <- jan_APC %>%
  group_by(`# Date`, Vehicle, Trip) %>%
  summarise(Boardings.tot = sum(Boardings),
            Deboardings.tot = sum(Deboardings),
            Max.Load = max(Load))
feb_APC.agg <- feb_APC %>%
  group_by(`# Date`, Vehicle, Trip) %>%
  summarise(Boardings.tot = sum(Boardings),
            Deboardings.tot = sum(Deboardings),
            Max.Load = max(Load))
mar_APC.agg <- mar_APC %>%
  group_by(`# Date`, Vehicle, Trip) %>%
  summarise(Boardings.tot = sum(Boardings),
            Deboardings.tot = sum(Deboardings),
            Max.Load = max(Load))
```

## Validate remaining trip codes used in APC and driver estimation data

Just checking which trips that We will have to manually attribute to driver estimations, such as trips with tripper service or service IDs that we don't know of, such as 21.

Much of January is not workable due to missing rows from our data cleaning and the inclusion of winter break, which is hard to attribute to driver estimations because it uses codes not currently existing in the GTFS. All rows that aren't working here will be excluded from the analysis.


```{r}
# jan_APC[!(jan_APC$Trip %in% trips$trip_id),]
# feb_APC[!(feb_APC$Trip %in% trips$trip_id),]
# mar_APC[!(mar_APC$Trip %in% trips$trip_id),]

# Display rows with invalid trip codes
# January
jan_DRE %>%
  filter(!(trip_id %in% trips$trip_id)) %>%
  filter(!grepl("_2", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_11", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_13", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_15", trip_id, fixed = TRUE)) %>%
  arrange(trip_id)

# February
feb_DRE %>%
  filter(!(trip_id %in% trips$trip_id)) %>%
  filter(!grepl("_2", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_11", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_13", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_15", trip_id, fixed = TRUE)) %>%
  arrange(trip_id)

# March
mar_DRE %>%
  filter(!(trip_id %in% trips$trip_id)) %>%
  filter(!grepl("_2", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_11", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_13", trip_id, fixed = TRUE)) %>%
  filter(!grepl("_15", trip_id, fixed = TRUE)) %>%
  arrange(trip_id)
```

## Attribute APC trips to driver estimation trips
```{r}
jan <- merge(jan_APC.agg, jan_DRE, by.x = c('# Date', 'Trip'), by.y = c('Date','trip_id'))
feb <- merge(feb_APC.agg, feb_DRE, by.x = c('# Date', 'Trip'), by.y = c('Date','trip_id'))
mar <- merge(mar_APC.agg, mar_DRE, by.x = c('# Date', 'Trip'), by.y = c('Date','trip_id'))

# Combine dataset because why not
combined <- rbind(mar, rbind(jan, feb))
```

## Analysis

Calculate differences between APC max load and passengers

```{r}
combined$Diff <- combined$Max.Load - combined$Passengers
```

Plotting

``` {r, echo=FALSE}
ggplot() +
  geom_histogram(data = combined,
                 aes(x = Diff),
                 binwidth = 1,
                 color="black",
                 fill="white") +
  geom_vline(aes(xintercept=mean(combined$Diff)),
            color="blue", linetype="dashed", line_width=2) +
  xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
  ylab("Number of Observations")
```

## Plots disaggregated by route

```{r, echo=FALSE}
for (route in unique(combined$Line...13)) {
  route_subset <- combined %>% filter(Line...13 == route)
  plot <- ggplot() +
    geom_histogram(data = route_subset,
                   aes(x = Diff),
                   binwidth = 5,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(route_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle(paste('Route', route)) +
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75))
  print(plot)
}
```

## Plots disaggregated by buses

```{r, echo=FALSE}
for (bus in unique(combined$Vehicle.x)) {
  bus_subset <- combined %>% filter(Vehicle.x == bus)
  plot <- ggplot() +
    geom_histogram(data = bus_subset,
                   aes(x = Diff),
                   binwidth = 5,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(bus_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle(paste('Bus', bus))+
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75))
  print(plot)
}
```

## Disaggregate by service ID

```{r, echo=FALSE}
# Make service ID column
combined$service_id <- 0
for (i in 1:nrow(combined)) {
  combined$service_id[i] <- as.numeric(strsplit(combined$Trip[i], split = '_', fixed=T)[[1]][2])
}

# Plot
for (service in unique(combined$service_id)) {
  service_subset <- combined %>% filter(service_id == service)
  plot <- ggplot() +
    geom_histogram(data = service_subset,
                   aes(x = Diff),
                   binwidth = 5,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(service_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle(paste('Service ID', service)) +
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75))
  print(plot)
}
```

## Disaggregate by AM vs. PM

```{r, echo=FALSE}
AM_subset <- combined %>% filter(as.numeric(Time) < 1200)
plot <- ggplot() +
    geom_histogram(data = AM_subset,
                   aes(x = Diff),
                   binwidth = 5,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(AM_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle('AM Observations') +
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75)) +
    scale_y_continuous(limits = c(0, 275),
                       breaks = c(0, 100, 200))
print(plot)

PM_subset <- combined %>% filter(as.numeric(Time) > 1200)
plot <- ggplot() +
    geom_histogram(data = PM_subset,
                   aes(x = Diff),
                   binwidth = 5,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(PM_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle('PM Observations') +
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75)) +
    scale_y_continuous(limits = c(0, 275),
                       breaks = c(0, 100, 200))
print(plot)
```

## Disaggregate by inbound vs. outbound

```{r}
outbound_subset <- combined %>% filter(Direction == 'outbound')
plot <- ggplot() +
    geom_histogram(data = outbound_subset,
                   aes(x = Diff),
                   binwidth = 1,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(outbound_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle('Outbound Observations') +
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75)) +
    scale_y_continuous(limits = c(0, 100),
                       breaks = c(0, 20, 40, 60, 80, 100))
print(plot)

inbound_subset <- combined %>% filter(Direction == 'inbound')
plot <- ggplot() +
    geom_histogram(data = inbound_subset,
                   aes(x = Diff),
                   binwidth = 1,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(inbound_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle('Inbound Observations') +
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75)) +
    scale_y_continuous(limits = c(0, 100),
                       breaks = c(0, 20, 40, 60, 80, 100))
print(plot)
```


## Disaggregate by driver estimate buckets

```{r}
# First make quintiles to disaggregate by
quintile <- quantile(combined$Passengers, probs = seq(0, 1, 0.2))


# First make histogram to decide buckets
for (i in 2:length(quintile)) {
  quintile_subset <- combined %>% filter(Passengers >= quintile[i-1] & Passengers <= quintile[i])
  plot <- ggplot() +
    geom_histogram(data = quintile_subset,
                   aes(x = Diff),
                   binwidth = 1,
                   color="black",
                   fill="white") + 
    geom_vline(aes(xintercept=mean(quintile_subset$Diff)),
               color="blue", linetype="dashed", size=1) +
    xlab("Difference in Max Passenger Load APC vs. Driver Estimation") +
    ylab("Number of Observations") +
    ggtitle(paste('Observations with max load between', quintile[i-1], 'and', quintile[i])) +
    scale_x_continuous(limits = c(-100, 75),
                       breaks = c(-100, -75, -50, -25, 0, 25, 50, 75)) +
    scale_y_continuous(limits = c(0, 50),
                       breaks = c(0, 10, 20, 30, 40, 50))
  print(plot)
}
```

